Designing an effective intervention of motivational interview using sequence classification methods

### 2017-02-27 paper/AMIA/main.tex
update introduction and conclusion sections

In this paper, our target area of application is motivational interview. In particular, this paper describes our work on the classification of utterance sequences. Classification of utterance sequences to distinguish different behavior outcomes such as positive change talk or commitment verses negative change talk or commitment, is an important part of clinical research aimed at designing effective interventions for many conditions and disorders. In this study, we focus on the transcripts of Motivational Interviews with obese adolescents (teens) and their caregivers. Childhood obesity is a serious public health concern in the United States and worldwide. Recent estimates indicate that approximately one third (31.8\%) of US children age 2-19 years are overweight and 16.9\% are obese \citep{01}. Adolescents who are obese are likely continue to be obese in adulthood and have a greater risk of heart disease, type 2 diabetes, stroke, cancer, and osteoarthritis \citep{02}. Therefore, there is a need for informatics-based methods to facilitate development of effective interventions for childhood obesity. One approach to designing intervention is Motivational Interviewing (MI), an evidence-based counseling technique to increase intrinsic motivation and self-efficacy for health-related behavior change \citep{03,04}. The goal of a MI counseling session is to encourage patients to explore their own desires, ability, reasons, need for and commitment to the targeted behavior change. These statements, referred to as ``change talk'' (or CT), consistently predict actual behavior change \citep{05} that can be sustained for as long as 34 months after an interview \citep{06}.

Our previous studies \cite{kotov2015interpretable, hasan2016study} explored several machine learning methods for the automatic annotation of clinical interview fragments, which uses a codebook \cite{carcone2013provider} with 115 different behavior codes. In this study, probabilistic models are employed for the analysis of sequential data to determine provider-patient communication sequences that are likely to translate into change talk and commitment language. Currently, a very similar work \cite{jaber2016multi} was done on education data for sequence classifications. Although Hidden Markov Model was originally applied to speech recognition \cite{rabiner1989tutorial}, this is one of the most successful method \cite{mutsam2016maximum, eickeler1998hidden, won2004training, chai2001folk} for sequence classification. Therefore, we utilize Markov Chain and Hidden Markov Model for the classification of utterance sequences.

\section*{Methods}
\subsection*{\textit{Data collection}}
The MYSCOPE \citep{07} codebook contains a total of 115 different codes that are grouped into the youth, caregiver, and counselor code groups. The experimental datasets for this work were constructed based on the transcripts of 37 motivational interview sessions, which
include a total of 11,353 segmented and annotated utterances. These utterances have been further partitioned into two subsets based on the structure of motivational interview
sessions: one dataset that includes all utterances from the adolescent sessions (6,579 samples) and the other dataset that includes all utterances from the caregiver sessions
(4,774 samples). A fragment of an adolescent session transcript is presented in Table~\ref{table:06}.

@@

\textbf {Markov Model}: In probability theory, a markov model \citep{45} is a stochastic model used to model randomly changing systems where it is assumed that future states depend only on the present state and not on the sequence of events that preceded it (that is, it assumes the markov property). Generally, this assumption enables reasoning and computation with the model that would otherwise be intractable. For the sequential analysis, we built two markov models $M$ and $\overline{M}$ describing provider strategies and patient responses in case of successful ($M$) and unsuccessful ($\overline{M}$) motivational interviews. A markov model $M$ can be represented as a weighted directed graph $G = (V, E, p)$, in which:

@@

\begin{figure}[htb!]
    \centering
    \includegraphics[width=0.70\textwidth]{figures/deltadata.jpg}
    \caption{\textbf{Characteristics of proposed markov model by varying the value of $\delta$}}
    \label{fig:delta}
\end{figure}

\subsection*{\textit{Evaluation methods}}
To ensure the robustness of performance estimates, we used 10-fold cross validation for the annotation of clinical interview fragments, and 5-fold cross validation for the classification of sequence of behavior codes \citep{36} as an experimental design. The performance of different classifiers and feature sets was evaluated in terms of precision, recall, F1 score (F1), kappa measure and accuracy using weighted macro-averaging over folds.

@@

Depending on the type of the interview transcript and the codebook size, SVM-AF achieves 3\%--9\% higher accuracy and 4\%--10\% higher F1 score than SVM and 4\%--10\% higher accuracy and 4\%--11\% higher F1 score than CRF, which highlights the importance of contextual and semantic features.

@@

@@

Two markov models are used to obtain top five most likely successful and unsuccessful motivational interviews that describing provider strategies and patient responses. Table~\ref{table:topsec} shows the top five successful and unsuccessful PPC code sequences.

@@

\section*{Acknowledgments}
We would like to thank the technical staff at Henry Ford Health Systems and Textual Data Analytics Laboratory at Wayne State University for their help with collecting and transforming binary data into free text EHR used for experiments reported in this paper.

>>>>

In this paper, our target area of application is motivational interview. In particular, this paper describes our work on the classification of utterance sequences. Classification of utterance sequences to distinguish different behavior outcomes such as positive change talk or commitment verses negative change talk or commitment, is an important part of clinical research aimed at designing effective interventions for many conditions and disorders. In this study, we focus on the transcripts of Motivational Interviews with obese adolescents (teens) and their caregivers. Childhood obesity is a serious public health concern in the United States and worldwide. Recent estimates indicate that approximately one third (31.8\%) of US children age 2-19 years are overweight and 16.9\% are obese \cite{ogden2012prevalence}. Adolescents who are obese are likely continue to be obese in adulthood and have a greater risk of heart disease, type 2 diabetes, stroke, cancer, and osteoarthritis \cite{general2010surgeon}. Therefore, there is a need for informatics-based methods to facilitate development of effective interventions for childhood obesity. One approach to designing an effective intervention is Motivational Interviewing (MI), an evidence-based counseling technique to increase intrinsic motivation and self-efficacy for health-related behavior change. The goal of a MI counseling session is to encourage patients to explore their own desires, ability, reasons, need for and commitment to the targeted behavior change. These statements, referred to as ``change talk'' (or CT), consistently predict actual behavior change that can be sustained for as long as 34 months after an interview.

Our previous studies \cite{kotov2015interpretable, hasan2016study} explored several machine learning methods for the automatic annotation of clinical interview fragments, which uses a codebook \cite{carcone2013provider} with 115 different behavior codes. In this study, probabilistic models are employed for the analysis of sequential data to determine provider-patient communication sequences that are likely to translate into change talk and commitment language. To the best of our knowledge, there have been very limited studies on sequence classification in motivational interview such as patient-provider dialogue. A sequential analysis \cite{eide2004physician} was performed on physician-patient dialogue to find the relationship between physician and patient behaviors associated with patients' exibiting cues. Currently, a very similar work \cite{jaber2016multi} was done on education data, where a sequence classification model was developed to predict the role of a student in a project team based on their communication patterns. Although Hidden Markov Model was originally applied to speech recognition \cite{rabiner1989tutorial}, this is one of the most successful method \cite{mutsam2016maximum, eickeler1998hidden, won2004training, chai2001folk} for sequence classification. Therefore, we utilize Markov Chain and Hidden Markov Model for the classification of utterance sequences.

\section*{Methods}
\subsection*{\textit{Data collection}}
The MYSCOPE \cite{} codebook contains a total of 115 different codes that are grouped into the youth, caregiver, and counselor code groups. The experimental datasets for this work were constructed based on the transcripts of 37 motivational interview sessions, which
include a total of 11,353 segmented and annotated utterances. These utterances have been further partitioned into two subsets based on the structure of motivational interview
sessions: one dataset that includes all utterances from the adolescent sessions (6,579 samples) and the other dataset that includes all utterances from the caregiver sessions
(4,774 samples). A fragment of an adolescent session transcript is presented in Table~\ref{table:06}. \\

@@

\textbf {Markov Model}: In probability theory, a markov model \cite{} is a stochastic model used to model randomly changing systems where it is assumed that future states depend only on the present state and not on the sequence of events that preceded it (that is, it assumes the markov property). Generally, this assumption enables reasoning and computation with the model that would otherwise be intractable. For the sequential analysis, we built two markov models $M$ and $\overline{M}$ describing provider strategies and patient responses in case of successful ($M$) and unsuccessful ($\overline{M}$) motivational interviews. A markov model $M$ can be represented as a weighted directed graph $G = (V, E, p)$, in which:

@@

\begin{figure}[htb!]
    \centering
    \includegraphics[width=0.70\textwidth]{figures/deltadata.png}
    \caption{\textbf{Characteristics of proposed markov model by varying the value of $\delta$}}
    \label{fig:delta}
\end{figure}

\subsection*{\textit{Evaluation methods}}
To ensure the robustness of performance estimates, we used 10-fold cross validation for the annotation of clinical interview fragments, and 5-fold cross validation for the classification of sequence of behavior codes \cite{} as an experimental design. The performance of different classifiers and feature sets was evaluated in terms of precision, recall, F1 score (F1), kappa measure and accuracy using weighted macro-averaging over folds.

@@

Depending on the type of the interview transcript and the codebook size, SVM-AF achieves 3\%--9\% higher accuracy and 4\%--10\% higher F1 score than SVM and 4\%--10\% higher accuracy and 4\%--11\% higher F1 score than CRF, which highlights the importance of contextual and semantic features. \\

@@

Two markov models are used to obtain top five most likely successful and unsuccessful motivational interviews that describing provider strategies and patient responses. Table~\ref{table:topsec} shows the top five successful and unsuccessful PPC code sequences. \\

@@

@@

Direction of future work include the exploration of other methods such as BLAST \cite{altschul1990basic} and Strand \cite{drew2014strand} for the sequence classification.

\section*{Acknowledgments}
We would like to thank the technical staff at Pediatric Prevention Research Center and Textual Data Analytics Laboratory at Wayne State University for their help with collecting and transforming data into free text EHR used for experiments reported in this paper.
